A Horrocks-type theorem for even orthogonal $K_2$

### 2019-08-27 dl-paper.tex
Minor fixes

Recall that to every irreducible root system $\Phi$ and a commutative ring $R$ one can associate two groups: the group $G(\Phi, R)$ called {\it simply-connected Chevalley group} (see~\cite[\S~3]{St71} or~\cite{VP}) and the group $\St(\Phi, R)$ called {\it Steinberg group} (see~\cref{sec:Steinberg-intro} for the definition). There is a well-defined map $\pi \colon \St(\Phi, R) \to G(\Phi, R)$ sending each generator $x_\alpha(\xi)$ to the elementary root unipotent $t_\alpha(\xi)$. The cokernel and the kernel of this map are, by definition, the $\K_1$ and $\K_2$ groups modeled on Chevalley group $G(\Phi, -)$:
\[ \xymatrix{ 1 \ar[r] & \K_2(\Phi, R) \ar[r] & \St(\Phi, R) \ar^{\pi}[r] & G(\Phi, R) \ar[r] & \K_1(\Phi, R) \ar[r] & 1.} \]

@@

Our main result is the following theorem, which is the orthogonal analogue of~\cite[Theorem~5.1]{Tu83} and the $\K_2$-analogue of~\cite[Theorem~6.8]{Su82} (cf. also with~\cite[Theorem~VI.5.2]{Lam10} and~\cite[Theorem~1.1]{St-poly}).
\begin{theorem}[Horrocks theorem for orthogonal $\K_2$]\label{thm:main} Let $A$ be a commutative ring in which $2$ is invertible. Then for any $\ell \geq 7$ the following commutative square is a pullback square in which all maps are injective:
\begin{equation} \nonumber
  \xymatrix{ \KO_2(2\ell, A) \ar[r] \ar[d] & \KO_2(2\ell, A[X]) \ar[d] \\ \KO_2(2\ell, A[X\inv]) \ar[r] & \KO_2(2\ell, A[X, X\inv]).}  
\end{equation} 
Moreover, the same assertion holds if one replaces the functor $\KO_2(2\ell, -)$ with $\K_2(\rD_\ell, -)$ or $\St(\rD_\ell, -)$. \end{theorem}
In the above statement $\KO_2(2\ell, -)$ denotes the unstable orthogonal $K_2$-functor (see~\cref{sec:quillen}).

@@

\subsection{Acknowledgements} \cref{thm41} was obtained by the first-named author who acknowledges the financial support of RSF grant No 17-11-01261.
\cref{thm:P1glueing} was obtained by the second-named author who acknowledges the financial support of JSC Gazpromneft and RFBR grant No 18-31-20044.

The authors would also to express their gratitude to A.~Stavrova for posing the problem and also to A.~Stepanov and N.~Vavilov for many useful comments and their interest in this work.

@@

\begin{comment}
Suppose for a moment that $\langle \alpha, \beta \rangle = -1$ and  $\langle \beta, \alpha \rangle = -1$ then
\[ \{s, t^{-1} \} = \{s,  t^{-1}\}_\alpha = \{t, s^{-1} \}_\beta^{-1} = \{s^{-1}, t\} \]
In particular, $\{s, s^{-1}\} = \{s, s^{-1}\}^{-1}$ 
\end{comment}
The classical Matsumoto theorem (see~\cite[Theorem~5.10]{Ma69}) allows one to compute the group $\K_2(\Phi, R)$ in the special case when $R=k$ is a field. More concretely, it asserts that $\K_2(\Phi, k) \cong \K_2^\mathrm{MW}(k)$ provided $\Phi$ is symplectic and  $\K_2(\Phi, k) \cong \K_2^\mathrm{M}(k)$ otherwise. In the following lemma we 
recall the computation of the group $\K_2(\Phi, R)$ in the case $R=k[X, X\inv]$.
\begin{lemma}[Hurrelbrink--Morita--Rehmann]\label{K2-laurent-field} Let $\Phi$ be a reduced irreducible root system of type $\neq \rG_2$ and $k$ be arbitrary field. Then there is a split exact sequence of abelian groups
\[ \xymatrix{ 0 \ar[r] & \K_2(\Phi, k) \ar[r] & \K_2(\Phi, k[X, X^{-1}]) \ar[r] & H(\Phi, k) \ar[r] & 0,}\text{ in which} \]
\[ H(\Phi, k) = \left\{\begin{array}{ll} \K_1^\mathrm{MW}(k)& \text{if $\Phi$ is symplectic,}\\ \K_1(k) \cong k^\times & \text{otherwise.}  \end{array}\right. \]  \end{lemma}
\begin{proof} Let us first consider the case of nonsymplectic $\Phi$, in which one can find a long root $\alpha\in \Phi$ in such a way that there is a commutative diagram of abelian groups
\[\xymatrixcolsep{5pc}\xymatrix{k^* \ar[r]^-{h} \ar[rrd]_{\{-, X\}} & \K_2(\Phi, k[X, X^{-1}]) \ar[r] & \K_2(\Phi, k(X)) \ar[d]^{\cong} \\
                                                                                      &                                 & \K_2^\mathrm{M}(k(X)),} \]
in which $h = \{ -, X \}_{\alpha}$ and the vertical map is an isomorphism by Matsumoto theorem.
Notice that the diagonal map is split by the obvious residue homomorphism and therefore is injective. This, in turn, implies that $h$ is also injective.
The assertion of the lemma now follows from~\cite[Satz~3]{Hur77} which asserts that for a nonsymplectic $\Phi$ holds $\K_2(\Phi, k[X, X\inv]) = \mathrm{Im}(h) \oplus \K_2(\Phi, k)$.

Consider now the case when $\Phi$ is symplectic. In this case the assertion of the lemma is just a reformulation of~\cite[Theorem~B]{MR91},
which asserts that for $\ell \geq 1$ one has $\K_2(\rC_\ell, k[X, X\inv] \cong \K_2(\rC_\ell, k) \oplus P(k)$, where
$P(k)$ is the set $k^* \times I_2(k)$ with the group structure given by
\[ (u, y) \cdot (v, z) = (uv, y + z - \langle\langle u, v\rangle\rangle).\]
Here $I^2(k)$ stands for the second power of the fundamental ideal $I(k)$ in the Witt ring $W(k)$ of $k$.
Recall that $\K_1^\mathrm{MW}(k)$ is isomorphic to the pullback of the diagram:
\[ \xymatrix{ \K_1^\mathrm{MW}(k) \ar[r] \ar[d] & I(k) \ar[d] \\ \K_1(k) \ar[r] & I(k)/I^2(k), } \]
in other words, it consists of pairs $[u, x]$ such that $x - \langle \langle u \rangle \rangle \in I^2(k)$.
It is easy to verify that the map $[u, x] \mapsto (u, \langle\langle u \rangle\rangle - x)$ defines an isomorphism of $\K_1^\mathrm{MW}(k)$ and $P(k)$. \end{proof}

@@

The following result, which is a special case of~\cite[Theorem~9.13]{Sch16} of M.~Schlichting, plays a key role in the proof of~\cref{thm41}.
\begin{externaltheorem}[Bass Fundamental Theorem]\label{bass-ft} Suppose that $R$ is a regular ring such that $2 \in R^*$, 
then for any $i\geq 1$, $k\in \ZZ/4\ZZ$ there is a natural split exact sequence of abelian groups \[ \xymatrix{ 0 \ar[r] & \GW_i^{[k]}(R) \ar[r] & \GW_i^{[k]}(R[X, X^{-1}]) \ar[r]  & \GW_{i-1}^{[k-1]}(R) \ar[r] & 0.} \] \end{externaltheorem}
%\begin{proof} This is a very special case of and the remark that follows it. \end{proof}
We will use the above theorem only in the special case $i=0$, in which it boils down to an earlier result of J.~Hornbostel, see~\cite[Corollary~5.3]{Ho05}.

@@

Whenever the coefficient $\xi$ of a Steinberg generator $g = x_\alpha(\xi)$ of $\St(\Phi, B)$ is a homogeneous element of $B$, i.\,e. $\xi \in B_d$ for some $d \in \mathbb{Z}$,
 we call the corresponding generator $g$ {\it homogeneous of degree $d$}.
It is not hard to show that $\St(\Phi, B)$ can be presented by the set of all homogeneous Steinberg generators modulo the following set of Steinberg relations
 between them (below $a, a' \in B_d$, $b\in R_e$ and $d,e \in \mathbb{Z}$): 
\begin{align}
&\,\,x_{\alpha}(a)\cdot x_{\alpha}(a') =  x_{\alpha}(a+a'),                        & \tag{R$1_d$} \\
&\,[x_{\alpha}(a),\,x_{\beta}(b)]= x_{\alpha+\beta}(N_{\alpha, \beta} \cdot ab),   & \alpha + \beta \in \Phi,\ \tag{R$2_{d,e}$} \\
&\,[x_{\alpha}(a),\,x_{\beta}(b)]= 1,                                              & \alpha - \beta \in \Phi,\ \tag{R$3^\angle_{d,e}$} \\
&\,[x_{\alpha}(a),\,x_{\beta}(b)]= 1,                                              & \alpha \perp \beta.\ \tag{R$3^\bot_{d,e}$}
\end{align}
By the degree of a Steinberg relation we mean the maximum of degrees of generators that appear in it.
For example, the degree of every relation of type $\text{R2}_{d,e}$ is $\max(d,e,d+e)$ 
 while the degree of a relation of type $\text{R3}^\bot_{d,e}$ or $\text{R3}^\angle_{d,e}$ is $\max(d,e)$.

@@

The following proposition is the main result of this subsection and also a direct generalization of~\cite[Lemma~3.3]{Tu83}.
\begin{prop} \label{lemma33} For $\Phi=\rA_{\geq 4}, \rD_{\geq 5}, \rE_{6,7,8}$ and $n \geq 1$ the map $i_n\colon \St^{\leq n}(\Phi, B) \to \St^{\leq n+1}(\Phi, B)$ induced by the natural embedding of generators, is an isomorphism. In particular, the obvious map $\St^{\leq 1}(\Phi, B) \to \St(\Phi, B)$ is an isomorphism. \end{prop}
\begin{proof}
 We need to construct a map $j_n$ which would be the inverse of $i_n$. 
 We start with a word map $\widetilde{j}_n^\Phi \colon F\langle \mathcal{X}^\Phi_{\leq n+1} \rangle \to \St^{\leq n}(\Phi, B)$ defined via
 \[ \widetilde{j}_n(x_\alpha(at^k)) = \begin{cases} x_\alpha(at^k), & k\leq n; \\
      [x_{\alpha - \beta} (N_{\alpha-\beta, \beta} \cdot at^{k-1}), x_{\beta}(t)], & k = n+1, \end{cases} \]
 where $\beta$ is an arbitrary root of $\Phi$ forming a sharp angle with $\alpha$.
 Standard argument (cf.~\cite[Proposition~1.1]{Re75} or~\cite[Proposition~3.2.2]{RS76}) shows that $\widetilde{j}_n$ does not depend on the choice of $\beta$.

>>>>

Recall that to every irreducible root system $\Phi$ and a commutative ring $R$ one can associate two groups: the group $G(\Phi, R)$ called {\it simply-connected Chevalley group} (see~\cite[\S~3]{St71} or~\cite{VP}) and the group $\St(\Phi, R)$ called {\it Steinberg group} (see~\cref{sec:Steinberg-intro} for the definition). There is a well-defined map $\pi \colon \St(\Phi, R) \to G(\Phi, R)$ sending each generator $x_\alpha(\xi)$ to the elementary root unipotent $t_\alpha(\xi)$. The cokernel and the kernel of this map are, by definition, the $\K_1$ and $\K_2$ groups modeled on Chevalley group $G(\Phi, -)$ (see~\cite{St78}).
\[ \xymatrix{ 1 \ar[r] & \K_2(\Phi, R) \ar[r] & \St(\Phi, R) \ar^{\pi}[r] & G(\Phi, R) \ar[r] & \K_1(\Phi, R) \ar[r] & 1.} \]

@@

Our main result is the following theorem, which is the orthogonal analogue of~\cite[Theorem~5.1]{Tu83} and the $\K_2$-analogue of~\cite[Theorem~6.8]{Su82} (cf. also with~\cite[Theorem~VI.5.2]{Lam10} and~\cite[Theorem~1.1]{St-poly}).
\begin{theorem}[Horrocks theorem for orthogonal $\K_2$]\label{thm:main} Let $A$ be a commutative ring in which $2$ is invertible. Then for any $\ell \geq 7$ the following commutative square is a pullback square in which all maps are injective:
\begin{equation} \nonumber
  \xymatrix{ \KO_2(2\ell, A) \ar[r] \ar[d] & \KO_2(2\ell, A[X]) \ar[d] \\ \KO_2(2\ell, A[X\inv]) \ar[r] & \KO_2(2\ell, A[X, X\inv]).}  
\end{equation} 
Moreover, the same assertion holds if one replaces the functor $\KO_2(2\ell, -)$ with $\K_2(\rD_\ell, -)$ or $\St(\rD_\ell, -)$. \end{theorem}
In the above statement $\KO_2(2\ell, -)$ denotes the unstable orthogonal $\K_2$-functor (see~\cref{sec:quillen}).

@@

\subsection{Acknowledgements} 
\cref{thm41} was obtained with support from RSF grant 19-71-30002,
 while \cref{thm:P1glueing} was obtained with support from RFBR grant No 18-31-20044.
The authors of this paper were also supported by ``Native towns'', a social investment program of PJSC ``Gazprom Neft''.

The authors would also to express their gratitude to A.~Stavrova for posing the problem and also to A.~Stepanov and N.~Vavilov for their useful comments and interest in this work.

@@

\begin{comment}
Suppose for a moment that $\langle \alpha, \beta \rangle = -1$ and  $\langle \beta, \alpha \rangle = -1$ then
\[ \{s, t^{-1} \} = \{s,  t^{-1}\}_\alpha = \{t, s^{-1} \}_\beta^{-1} = \{s^{-1}, t\} \]
In particular, $\{s, s^{-1}\} = \{s, s^{-1}\}^{-1}$ 
\end{comment}
Recall that the classical Matsumoto theorem (see~\cite[Theorem~5.10]{Ma69}) allows one to compute the group $\K_2(\Phi, R)$ in the special case when $R=k$ is a field.
Using the modern language of Milnor--Witt K-theory (see~\cite{Mo04}) it can be formulated as the following computation:
\[\K_2(\Phi, k) = \left\{\begin{array}{ll} \K_2^\mathrm{MW}(k)& \text{if $\Phi$ is symplectic,}\\ \K_2^\mathrm{M}(k) & \text{otherwise.}\end{array}\right. \]
In the following lemma we recall the computation of the group $\K_2(\Phi, R)$ in the case $R=k[X, X\inv]$.
\begin{lemma}[Hurrelbrink--Morita--Rehmann]\label{K2-laurent-field} Let $\Phi$ be a reduced irreducible root system of type $\neq \rG_2$ and $k$ be arbitrary field. Then there is a split exact sequence of abelian groups
\[ \xymatrix{ 0 \ar[r] & \K_2(\Phi, k) \ar[r] & \K_2(\Phi, k[X, X^{-1}]) \ar[r] & H(\Phi, k) \ar[r] & 0,}\text{ in which} \]
\[ H(\Phi, k) = \left\{\begin{array}{ll} \K_1^\mathrm{MW}(k)& \text{if $\Phi$ is symplectic,}\\ \K_1(k) \cong k^\times & \text{otherwise.}  \end{array}\right. \]  \end{lemma}
\begin{proof} Let us first consider the case of nonsymplectic $\Phi$, in which one can find a long root $\alpha\in \Phi$ in such a way that there is a commutative diagram of abelian groups
\[\xymatrixcolsep{5pc}\xymatrix{k^* \ar[r]^-{h} \ar[rrd]_{\{-, X\}} & \K_2(\Phi, k[X, X^{-1}]) \ar[r] & \K_2(\Phi, k(X)) \ar[d]^{\cong} \\
                                                                                      &                                 & \K_2^\mathrm{M}(k(X)),} \]
in which $h = \{ -, X \}_{\alpha}$ and the vertical map is an isomorphism by Matsumoto theorem.
Notice that the diagonal map is split by the obvious residue homomorphism and therefore is injective. This, in turn, implies that $h$ is also injective.
The assertion of the lemma now follows from~\cite[Satz~3]{Hur77} which asserts that for a nonsymplectic $\Phi$ holds $\K_2(\Phi, k[X, X\inv]) = \mathrm{Im}(h) \oplus \K_2(\Phi, k)$.

Consider now the case when $\Phi$ is symplectic. In this case the assertion of the lemma is just a reformulation of~\cite[Theorem~B]{MR91},
which asserts that for $\ell \geq 1$ one has $\K_2(\rC_\ell, k[X, X\inv] \cong \K_2(\rC_\ell, k) \oplus P(k)$, where
$P(k)$ is the set $k^* \times I_2(k)$ with the group structure given by
\[ (u, y) \cdot (v, z) = (uv, y + z - \langle\langle u, v\rangle\rangle).\]
Here $I^2(k)$ stands for the second power of the fundamental ideal $I(k)$ in the Witt ring $W(k)$ of $k$.
Recall from~\cite{Mo04} that $\K_1^\mathrm{MW}(k)$ is isomorphic to the pullback of the diagram:
\[ \xymatrix{ \K_1^\mathrm{MW}(k) \ar[r] \ar[d] & I(k) \ar[d] \\ \K_1(k) \ar[r] & I(k)/I^2(k), } \]
in other words, it consists of pairs $[u, x]$ such that $x - \langle \langle u \rangle \rangle \in I^2(k)$.
It is easy to verify that the map $[u, x] \mapsto (u, \langle\langle u \rangle\rangle - x)$ defines an isomorphism of $\K_1^\mathrm{MW}(k)$ and $P(k)$. \end{proof}

@@

The following result, which is a special case of~\cite[Theorem~9.13]{Sch16} of M.~Schlichting, plays a key role in the proof of~\cref{thm41}.
\begin{externaltheorem}[Bass Fundamental Theorem]\label{bass-ft} Suppose that $R$ is a regular ring such that $2 \in R^*$, 
then for any $i\geq 1$, $k\in \ZZ/4\ZZ$ there is a natural split exact sequence of abelian groups \[ \xymatrix{ 0 \ar[r] & \GW_i^{[k]}(R) \ar[r] & \GW_i^{[k]}(R[X, X^{-1}]) \ar[r]  & \GW_{i-1}^{[k-1]}(R) \ar[r] & 0.} \] \end{externaltheorem}
%\begin{proof} This is a very special case of and the remark that follows it. \end{proof}
We will use the above theorem only in the special case $k=0$, in which it boils down to an earlier result of J.~Hornbostel, see~\cite[Corollary~5.3]{Ho05}.

@@

Whenever the coefficient $\xi$ of a Steinberg generator $g = x_\alpha(\xi)$ of $\St(\Phi, B)$ is a homogeneous element of $B$, i.\,e. $\xi \in B_d$ for some $d \in \mathbb{Z}$,
 we call the corresponding generator $g$ {\it homogeneous of degree $d$}.
It is not hard to show that $\St(\Phi, B)$ can be presented by the set of all homogeneous Steinberg generators modulo the following set of Steinberg relations
 between them (below $a, a' \in B_d$, $b\in R_e$ and $d,e \in \mathbb{Z}$): 
\begin{align}
&\,\,x_{\alpha}(a)\cdot x_{\alpha}(a') =  x_{\alpha}(a+a'),                        & \tag{R$1_d$} \\
&\,[x_{\alpha}(a),\,x_{\beta}(b)]= x_{\alpha+\beta}(N_{\alpha, \beta} \cdot ab),   & \alpha + \beta \in \Phi,\ \tag{R$2_{d,e}$} \\
&\,[x_{\alpha}(a),\,x_{\beta}(b)]= 1,                                              & \alpha - \beta \in \Phi,\ \tag{R$3^\angle_{d,e}$} \\
&\,[x_{\alpha}(a),\,x_{\beta}(b)]= 1,                                              & \alpha \perp \beta.\ \tag{R$3^\bot_{d,e}$}
\end{align}
By the degree of a Steinberg relation we mean the maximum of degrees of generators that appear in the relation.
For example, the degree of every relation of type $\text{R2}_{d,e}$ is $\max(d,e,d+e)$ 
 while the degree of a relation of type $\text{R3}^\bot_{d,e}$ or $\text{R3}^\angle_{d,e}$ is $\max(d,e)$.

@@

The following proposition is the main result of this subsection and also a direct generalization of~\cite[Lemma~3.3]{Tu83}.
\begin{prop} \label{lemma33} For $\Phi=\rA_{\geq 4}, \rD_{\geq 5}, \rE_{6,7,8}$ and $n \geq 1$ the map $i_n\colon \St^{\leq n}(\Phi, B) \to \St^{\leq n+1}(\Phi, B)$ induced by the natural embedding of generators, is an isomorphism. In particular, the obvious map $\St^{\leq 1}(\Phi, B) \to \St(\Phi, B)$ is an isomorphism. \end{prop}
\begin{proof}
 We need to construct a map $j_n$ which would be the inverse of $i_n$. 
 We start with a word map $\widetilde{j}_n^\Phi \colon F\langle \mathcal{X}^\Phi_{\leq n+1} \rangle \to \St^{\leq n}(\Phi, B)$ defined via
 \[ \widetilde{j}_n(x_\alpha(at^k)) = \begin{cases} x_\alpha(at^k), & k\leq n; \\
      [x_{\alpha - \beta} (N_{\alpha-\beta, \beta} \cdot at^{k-1}), x_{\beta}(t)], & k = n+1, \end{cases} \]
 where $\beta$ is any root of $\Phi$ forming a sharp angle with $\alpha$.
 Standard argument (cf.~\cite[Proposition~1.1]{Re75} or~\cite[Proposition~3.2.2]{RS76}) shows that $\widetilde{j}_n$ does not depend on the choice of $\beta$.
